New seismological constraints on the available parameter space of macroscopic dark matter

### 2016-09-27 LaTeX/Current/Article_Draft_EDITED.tex
More Update
Small correction
corrected again
Updated Article_Draft_EDITED.pdf

\section{Simulation}
In previous works, e.g. \cite{banerdt2005seismic}, the homogeneous Earth and Moon models have been used to place constraints of the parameter space of macroscopic dark matter. This approach neglects lensing which occurs because of the velocity gradient and the spherical boundaries of strata, and neglects the energy losses from reflection and refraction across these boundaries. We use the analysis in section III to propagate rays, each carrying a fraction of the total energy deposited by the Macro, to the boundary of the Moon. We then use HEALPix to create an intensity map of the lunar surface for a representative sample of Macro impacts. We then convert the intensity maps to displacement maps, and using the sensitivity of the lunar seismometers, we obtain an average event rate that the lunar seismometers would measure.
\subsection{Data Generation}
We consider a Macro trajectory which passes a distance $D$ from the center of the moon. The Macro trajectory has length $L$, and along its trajectory we consider a set of $M$ points. From each of these points we propagate $N$ randomly oriented rays, each endowed with $E_i = \rho(r_{i})\sigma_Xv_X^2 L/(MN)$ energy, where $i\in\{1,\dots,M\}$ labels the point, and $\rho$ is taken to be a function of the radius. Because of the spherical symmetry, each ray is confined to a subspace of two dimensions. We use the trajectories in section III to propagate the rays. Since the moon is stratified, we must halt the propagation at the boundary of each layer to reflect and transmit rays. During a ray's propagation through a given stratum, its time spent in that layer $\Delta t_i$ is recorded, the attenuation factor in that layer $\te{Att}_i$ is recorded, and the reflection/refraction coefficient $\te{Ref}_i$ is recorded. Each time a ray makes contact with the surface of the Moon, its position, $\te{Att}$, $\te{Ref}$, and $\Delta t$ are recorded. The data produced by the propagation routine is a table of these values and the corresponding surface coordinates, as well as density at the source, $\kappa$ at the source, and $p$-wave velocity at the source. In this model, it takes 16 iterations to propagate a ray from one side of the moon to the other. Rays which propagate from one side of the moon to the other only after 16 iterations will be weak from undergoing 16 refractions. Thus in order to save on computation time, it is reasonable to reduce the number of iterations to 9.
\subsection{Data Analysis}
Analysis begins by converting the position data to HEALPix coordinates for a given $n_\te{side}$. We pick $n_\te{side} = 4$ since this corresponds to an angular resolution of 7 degrees, which accounts for the effects of diffraction about the strata boundaries. Define $A_\te{pix} = 4\pi R_\te{moon}^2/n_\te{pix}$ the area of a HEALPix pixel, where $n_\te{pix} = 12\times 2^{n_{\te{side}}}$. The time it takes a ray to cross a HEALPix pixel can be estimated $t_\te{crossing}^2 = A_\te{pix}/\pi v_\te{surface}^2$. According to VPREMOON, $v_\te{surface} = 1\,\te{km s}^{-1}$. Rays that contact a pixel within $t_\te{crossing}$ of one another are considered to add constructively. Since the integral in section IV is unreasonable to do for more than one ray, we average $\te{Att}$ over the coincident rays. We require that each ray distributes its energy evenly over the area of the HEALPix pixel, and thus set $A = A_\te{pix}$. This is a good approximation when $N=n_\te{pix}$, since the average angular separation of the rays at the source corresponds to the average angular separation of the HEALPix pixels.

@@

This procedure is repeated for 17 values of $D$ spaced evenly by $100\,\te{km}$. The average fraction of detectable surface area is then approximately
\begin{align}
\scr F =\f{\sum_D \scr F_D}{\sum_D D}\,.
\end{align}
If there are $n$ lunar seismometers, we suppose that each occupies its own HEALPix pixel. Define $m = \scr Fn_\te{pix}$, the average number of pixels with detectable frequency. If $n+m\geq n_\te{pix}$, then detection is guaranteed. Otherwise, if $\scr M$ Macros impact the moon, the likelihood of any of them being detected is
\begin{align}
\scr P &= 1-\ps{\f{(n_\te{pix}-n)!(n_\te{pix}-m)!}{n_\te{pix}!(n_\te{pix}-n-m)!}}^{\scr M}\,.
\end{align}

Since it is unlikely that a Macro impact will be detected if it only just exceeds one digital unit of displacement, it is useful to use a weighted sum. As the strength of the signal goes to $\infty$, we should always expect impact to be seen, and as the displacement approaches $d = 3\times 10^{-10}\te{ m}$, we should expect no chance of detection. Hence, a tile with displacement $\Delta x$ has detection likelihood
\begin{align}
1-e^{p(1-\f{\Delta x}{d})}\,,
\end{align}
where $p$ describes how the detection probability increases as the strength of signal increases.
\subsection{Homogeneous Limit}
To obtain an estimate of the size of a detectable Macro, we suppose that one ray from each point source on the Macro Trajectory hits a single pixel. In the case that rays don't bend, this is a good approximation for behavior away from the source. Note, however, that this does not account for geometric focusing e.g. around the entrance and exit of the Macro trajectory.
\begin{align}
u\leq\sqrt{\f{\Xi E}{\pi^2\rho k v_p^2 R_\te{Moon}^2}}\,.
\end{align}
This expression is independent of the choice of $n_\te{side}$. For the triangle wave, $\sigma_X\leq 1\times 10^{-7}\te{\,cm}^2$ is unlikely to be detected unless it enters or exits the Moon adjacent to a seismometer.
\section{Results}
Data has been generated but not run through the analysis routine yet. Preliminary results suggest that the computer analysis will not produce results much stronger than (30).
\section{Conclusion}
For Macros with density sufficient to pass nearly unimpeded through the Moon and sufficiently small cross-section, and in particular nuclear dense objects such as strange quark nuggets with cross-section $\sigma_X\leq 1\times 10^{-7}\te{\, cm}^2$, we have demonstrated that seismological constraints are insufficient to rule them out. Since the Moon (and Earth) preferentially attenuate higher frequency modes, it is unlikely that measuring high frequencies would greatly improve sensitivity to these types of impacts. On the other hand, increasing the lunar seismometer sensitivity to small displacements may allow one to use seismological evidence to rule out high density Macros, although it is uncertain whether seismic noise at such small displacements would make Macro impacts indistinguishable from the background.

>>>>

\section{Simulation}
In previous works, e.g. \cite{banerdt2005seismic}, the homogeneous Earth and Moon models have been used to place constraints of the parameter space of macroscopic dark matter. This approach neglects lensing which occurs because of the velocity gradient and the spherical boundaries of strata, and neglects the energy losses from reflection and refraction across these boundaries. We use the analysis in section III to propagate rays, each carrying a fraction of the total energy deposited by the Macro, to the boundary of the Moon. We then use HEALPix to create an intensity map of the lunar surface for a representative sample of Macro impacts. Finally, we convert the intensity maps to displacement maps, and using the sensitivity of the lunar seismometers, we obtain an average event rate that the lunar seismometers would measure.
\subsection{Data Generation}
We consider a Macro trajectory which passes a distance $D$ from the center of the moon. The Macro trajectory has length $L$ within the moon, and along its trajectory we consider a set of $M$ points. From each of these points we propagate $N$ randomly oriented rays, each endowed with $E_i = \rho(r_{i})\sigma_Xv_X^2 L/(MN)$ energy, where $i\in\{1,\dots,M\}$ labels the point, and $\rho$ is taken to be a function of the radius. Because of the spherical symmetry, each ray is confined to a subspace of two dimensions. We use the trajectories in section III to propagate the rays. Since the moon is stratified, we must halt the propagation at the boundary of each layer to reflect and transmit rays. During a ray's propagation through a given stratum, its time spent in that layer $\Delta t_i$, the attenuation factor in that layer $\te{Att}_i$, and the reflection/refraction coefficient $\te{Ref}_i$ are all recorded. Each time a ray makes contact with the surface of the Moon, its position, $\te{Att}$, $\te{Ref}$, and $\Delta t$ are recorded. The data produced by the propagation routine is a table of these values and the corresponding surface coordinates (i.e. we only count the energy of those rays which hit the surface), as well as density at the source, $\kappa$ at the source, and $p$-wave velocity at the source. In this model, it takes 16 iterations to propagate a ray from one side of the moon to the other. However, we simplify the model by assuming constant $Q = 6750$. In this case, it takes 8 iterations to propagate entirely through the Moon.
\subsection{Data Analysis}
Analysis begins by converting the surface-incident ray position data to HEALPix coordinates for a given $n_\te{side}$. Define $A_\te{pix} = 4\pi R_\te{moon}^2/n_\te{pix}$ the area of a HEALPix pixel, where $n_\te{pix} = 12\times 2^{n_{\te{side}}}$. The time it takes a ray to cross a HEALPix pixel can be estimated $t_\te{crossing}^2 = A_\te{pix}/\pi v_\te{surface}^2$. According to VPREMOON, $v_\te{surface} = 1\,\te{km s}^{-1}$. Rays that contact a pixel within $t_\te{crossing}$ of one another are considered to add constructively. We require that each ray distributes its energy evenly over the area of the HEALPix pixel, and thus set $A = A_\te{pix}$. This is a good approximation when $N=n_\te{pix}$, since the average angular separation of the rays at the source corresponds to the average angular separation of the HEALPix pixels. Note, for the case of the Moon, we omit the effects of anelastic attenuation since its contribution is relatively unimportant, as indicated by several tests in which its average contribution was on the order of $e^{-10^{-2}}$.

@@

Since it is unlikely that a Macro impact will be detected if it only just exceeds one digital unit of displacement, it is useful to use a weighted sum. As the strength of the signal goes to $\infty$, we should always expect impact to be seen, and as the displacement approaches $d = 3\times 10^{-10}\te{ m}$, we should expect no chance of detection. Hence, a tile with displacement $\Delta x$ has detection likelihood
\begin{align}
1-\exp\ps{p\p{1-\f{\Delta x}{d}}}\,,
\end{align}
where $p$ describes how the detection probability increases as the strength of signal increases. Requiring that 4 digital units be $q$ times as detectable as 2 digital unit fixes $p = \ln\sqrt q$. Summing over the all the pixels yields the effective number of pixels in which there exists a detectable signal
\begin{align}
m_D = \sum_{i = 1}^{n_{\te{pix}}}\pc{1-\exp\ps{p\p{1-\f{\Delta x_i}{d}}}}
\end{align}

This procedure is repeated for 20 values of $D$ spaced evenly by $85\,\te{km}$. The average number of pixels on which there exists a detectable signal is then
\begin{align}
m =\f{\sum_D D m_D}{\sum_D D}\,.
\end{align}
If there are $n$ lunar seismometers (or tight clusters thereof), we suppose that each occupies one HEALPix pixel. If $n+m\geq n_\te{pix}$, then detection is guaranteed. Otherwise, if $\scr M$ Macros impact the moon, the likelihood of any of them being detected is
\begin{align}
\scr P &= 1-\ps{\f{(n_\te{pix}-n)!(n_\te{pix}-m)!}{n_\te{pix}!(n_\te{pix}-n-m)!}}^{\scr M}\,.
\end{align}
\subsection{Homogeneous Limit}
To obtain an estimate of the size of a detectable Macro, we suppose that one ray from each point source on the Macro Trajectory hits a single pixel. In the case that rays don't bend, this is a good approximation for behavior away from the source. Note, however, that this does not account for geometric focusing e.g. around the entrance and exit of the Macro trajectory.
\begin{align}
u\leq\sqrt{\f{\Xi E}{\pi^2\rho k v_p^2 R_\te{Moon}^2}}\,.
\end{align}
This expression is independent of the choice of $n_\te{side}$. For the triangle wave, $\sigma_X\leq 1\times 10^{-7}\te{\,cm}^2$ is unlikely to be detected unless it enters or exits the Moon adjacent to a seismometer.
\section{Results}
Data has been generated but not run through the analysis routine yet. Preliminary results suggest that the computer analysis will not produce results much stronger than (32).
\section{Conclusion}
For Macros with density sufficient to pass nearly unimpeded through the Moon and sufficiently small cross-section, and in particular nuclear dense objects such as strange quark nuggets with cross-section $\sigma_X\leq 1\times 10^{-7}\te{\, cm}^2$, we have demonstrated that seismological constraints are insufficient to rule them out. Since the Moon (and Earth) preferentially attenuate higher frequency modes, it is unlikely that measuring high frequencies would greatly improve sensitivity to these types of impacts. On the other hand, increasing the lunar seismometer sensitivity to small displacements may allow one to use seismological evidence to rule out high density Macros, although it is uncertain whether seismic noise at such small displacements would make Macro impacts indistinguishable from the background.
